Stochastic blockmodeling of relational event dynamics

### 2012-05-31 docs/abstract.tex
Paper edits.  Intro reorg.

\begin{abstract}
Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
For continuous-time network data, a number of approaches have recently been proposed for modeling the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
For static networks, methods such as stochastic blockmodels have traditionally been used for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
The method is illustrated using dyadic interaction data such as email communication and Twitter direct messages.
Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.
Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

>>>>

\begin{abstract}
Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
For continuous-time network data, a number of approaches have recently been proposed for modeling the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
For static networks, methods such as stochastic blockmodels have traditionally been used for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
The method is illustrated using dyadic interaction data such as email communication and Twitter direct messages.
Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
The fitted models have better predictive accuracy than baselines and
relational event models without the latent structure and a standard
stochastic blockmodel approach.
%Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

### 2012-05-31 docs/discussion.tex
Paper edits.  Intro reorg.

 We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).
This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.

%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.

%For example, events among nodes in cluster 1 have an increased propensity for reciprocal events -- after node A sends to node B, an event from B to A has an intensity that is $e^4$ larger, all else held constant.  Interestingly, degree effects seem to vary by block: in block (2,2) active senders are more likely to be a sender whereas in block (1,1) the number of previous occurrences for a given dyad has a larger multiplicative effect.

This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.
In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
We also note that we improve over the situation where we constrain $K=1$.
Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.

>>>>

 We propose a hierarchical approach for modeling event-based network data as a continuous-time Markov process.
Given a specification for the, the proposed method employs latent variables to model the heterogeneity in the event dynamics.
%is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
%The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).
%This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in the dynamics of the network as a whole.

This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.
In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
We also note that we can improve over relational event models that lack latent clusters as well as stochastic blockmodels for count data, both special cases 
Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.

### 2012-05-31 docs/intro.tex
Add figure for intro example.
Paper edits.  Intro reorg.

Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010} are a class of statistical models for static network data that employ latent variables to model unobserved heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.  % Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels

Network data, however, is often collected as a sequence of events occurring over time.
Recent work leverages  continuous-time models from event history analysis  \cite{AalenOddO.2008} to model network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}.
These models allow one to specify how the process depends on the previous history of events.
In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.

A limitation of all of these approaches is the assumption of a single common behavior for all individuals.
In contrast, in real-world networks it is reasonable to expect {\it heterogeneity} in dynamic behavior.  
In an organization, the manner in which an individual communicates with others may be a function of an individual's {\it role} in the organization.
For example, in a university,  email communication patterns over time between professors, students, and staff will likely be quite different within and between the three groups.
%PS: not totally happy with this example since the "roles" are completely determined by one's "job title" and I think we would prefer something
% less explicit, e.g., that there are "leaders", "followers", etc, that go beyond just job titles. So feel free to edit this further!

@@

% TODO: Incorporate notes from joint meeting:
% - used to thinking about actors differentiated by rates
% - other things in the dynamics than rates, e.g. form of interaction.
% - Also important aspect of modeling social dynamics.
% especially important in cases like plot 1(c) where classical methods
% will fail.

\begin{figure}

\caption{}
\label{fig:example}
\end{figure}

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data to capture node-level heterogeneity, where we assume that latent clusters of nodes share similar patterns of interaction.
We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model  with simulated data.
Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, allowing identification of particular \emph{roles} in communication.

>>>>

Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO] that is associated with some other quantity, e.g. a covariate such as age or sex, or perhaps shared [TODO].
Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.

If we are interested in studying the events in a network, there are other aspects of the dynamics in addition to the mean rate. 
For example, human conversation often has an increased propensity for reciprocated events.
Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
Figure \ref{fig:introexample} shows how an analysis using a blockmodels will be unable to learn about these differences.

%A limitation of all of these approaches is the assumption of a single common behavior for all individuals.
%In contrast, in real-world networks it is reasonable to expect {\it heterogeneity} in dynamic behavior.  
%In an organization, the manner in which an individual communicates with others may be a function of an individual's {\it role} in the organization.
%For example, in a university,  email communication patterns over time between professors, students, and staff will likely be quite different within and between the three groups.

% - used to thinking about actors differentiated by rates
% - other things in the dynamics than rates, e.g. form of interaction.
% - Also important aspect of modeling social dynamics.
% especially important in cases like plot 1(c) where classical methods
% will fail.

%PS: not totally happy with this example since the "roles" are completely determined by one's "job title" and I think we would prefer something
% less explicit, e.g., that there are "leaders", "followers", etc, that go beyond just job titles. So feel free to edit this further!

@@

Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
Our approach employs a flexible framework for  specifing how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
%extend continuous-time models from event history analysis to model sequences of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
%These models allow one to specify how the process depends on the previous history of events. % TODO use 'flexible'?
In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.

We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model  with simulated data.
Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, enabling the study of a dynamic extension to stochastic equivalence.
%allowing identification of particular \emph{roles} in  that enable

\begin{figure}
\centering
\includegraphics[scale=.8]{../figs/introexample/all}
\caption{}
\label{fig:example}
\end{figure}

### 2012-05-31 docs/model.tex
Paper edits.  Intro reorg.

 In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Appendix we show how this assumption also reduces the computational complexity of computing the above likelihood.\footnote{This is analogous to a Markov condition for this class of models.}

>>>>

 In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we allow each intensity function $\lambda_{ij}(t)$ to only change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
%\footnote{TODO: This is analogous to a Markov condition for this class of models.}

### 2012-05-31 docs/submission.tex
Paper edits.  Intro reorg.

\title{Stochastic Blockmodels for Relational Event Data}
% Latent Variable Modeling of Relational Event Dynamics
% Latent Variable Modeling of Relational Event Sequences
% Latent Variable Modeling for Relational Event Sequences

\author{
David S.~Hippocampus\thanks{ Use footnote for providing further information
about author (webpage, alternative address)---\emph{not} for acknowledging
funding agencies.} \\
Department of Computer Science\\
Cranberry-Lemon University\\
Pittsburgh, PA 15213 \\
\texttt{hippo@cs.cranberry-lemon.edu} \\
\And
Coauthor \\
Affiliation \\
Address \\
\texttt{email} \\
\AND
Coauthor \\
Affiliation \\
Address \\
}

>>>>

\title{Stochastic blockmodeling of relational event dynamics}
% Dynamic stochastic equivalence for relational events
% Stochastic Blockmodels for Relational Event Data}
% Latent Variable Modeling of Relational Event Dynamics
% Latent Variable Modeling of Relational Event Sequences
% Latent Variable Modeling for Relational Event Sequences

% \author{
% Christopher DuBois
% Department of Statistics\\
% University of California, Irvine\\
% \texttt{duboisc@ics.uci.edu} \\
% \And
% Coauthor \\
% Affiliation \\
% Address \\

% \AND
% Coauthor \\
% Affiliation \\
% Address \\
% }
